What Boolean Algebras Can Teach Us About Weighted Model Counting

### 2020-06-04 paper.tex
ADDMC

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{fullpage}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage[capitalise]{cleveref}
\usepackage{bm}
\usepackage{booktabs}
\usepackage{tikz}
\usepackage{xcolor}
\usepackage[backgroundcolor=lightgray]{todonotes}

@@

\usetikzlibrary{cd}

@@

>>>>

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{fullpage}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage[capitalise]{cleveref}
\usepackage{bm}
\usepackage{booktabs}
\usepackage{tikz}
\usepackage{xcolor}
\usepackage[backgroundcolor=lightgray]{todonotes}
\usepackage{complexity}

@@

\usetikzlibrary{cd}
\usetikzlibrary{bayesnet}
\usetikzlibrary{calc}

@@

\section{How to represent a Bayesian network}

Let $G = (V, E)$ be a DAG, and let $(X_v)_{v \in V}$ be the discrete random
variables (RVs). Let $(d_v)_{v \in V}$ denote the numbers of values that each RV
can take and assume that $d_v$ is a power of two for all $v \in V$.
\begin{enumerate}
\item For each $v \in V$, let $A_v$ be a set of $\log_2 d_v$ Boolean variables.
  (If $d_v$ is not a power of two, we would need to manually forbid each illegal
  value, e.g., by adding formulas such as $\neg(x_1 \land x_2 \land x_3 \land
  \neg x_4)$ to the theory. Alternatively, we could set their probability to
  zero, but the former should be better.)
\item For $v \in V$:
  \begin{enumerate}
  \item If $v$ has no parents, we can directly provide a probability measure on
    $\mathcal{F}(A_v)$.
  \item Otherwise, we need a probability measure on
    \[
      \mathcal{F} \left( A_v \cup \bigcup_{(u, v) \in E} A_u \right).
    \]
    For any element $a_u \land a_v$ in this free Boolean algebra such that $a_v
    \in A_v \setminus \{ 0, 1 \}$, and $a_u \not\in A_v$, we can set
    \[
      m(a_u \land a_v) = \Pr(a_v \mid a_u)\Pr(a_u).
    \]
  \end{enumerate}
\item When a \#\SAT{} algorithm asks for the weight of a model $\bigwedge_{v \in
    V} l_v$, split it into segments that correspond to each leaf, take the
  probabilities from each such node, and multiply them. NB: The segments will
  intersect, e.g., with
  \[
    \begin{tikzpicture}
      \node[latent] (C) {$C$};
      \node[latent, below left = of C] (A) {$A$};
      \node[latent, below right = of C] (B) {$B$};
      \node[latent, right = 2 of C] (D) {$D$};
      \edge {C} {A, B};
    \end{tikzpicture}
  \]
  $a \land b \land c \land d$ will be split into $(a \land c) \land (b \land c)
  \land d$.
\end{enumerate}

\subsection{Bayesian networks as Boolean algebras}

\begin{table}
  \caption{Examples of how Bayesian networks encode as Boolean algebras. Here,
    $A$ denotes $\mathcal{F}\{a\}$ and $AB$ denotes $\mathcal{F}\{a, b\}$.}
  \centering
  \begin{tabular}{cc}
    \toprule
    Bayesian network & Boolean algebra \\
    \midrule \\[-1.3em]
    \begin{tikzpicture}[baseline={(A.base)}]
      \node[latent] (A) {$A$};
    \end{tikzpicture} & $A$ \\
    \midrule \\[-1.3em]
    \begin{tikzpicture}[baseline={(A.base)}]
      \node[latent] (A) {$A$};
      \node[latent,right =of A] (B) {$B$};
    \end{tikzpicture} & $A + B$ \\
    \midrule \\[-1.3em]
    \begin{tikzpicture}[baseline={($(A.base)!.5!(B.base)$)}]
      \node[latent] (A) {$A$};
      \node[latent,below =of A] (B) {$B$};
      \edge {A} {B};
    \end{tikzpicture} & $AB$ \\
    \midrule \\[-1.3em]
    \begin{tikzpicture}[baseline={($(A.base)!.5!(B.base)$)}]
      \node[latent] (A) {$A$};
      \node[latent,below =of A] (B) {$B$};
      \node[latent,right =of A] (C) {$C$};
      \edge {A} {B};
    \end{tikzpicture} & $B +_A C$ \\
    \midrule \\[-1.3em]
    \begin{tikzpicture}[baseline={($(A.base)!.5!(B.base)$)}]
      \node[latent] (A) {$A$};
      \node[latent,below =of A] (B) {$B$};
      \node[latent,right =of A] (C) {$C$};
      \node[latent,below =of C] (D) {$D$};
      \edge {A} {B};
      \edge {C} {D};
    \end{tikzpicture} & $B +_{A+C} D$ \\
    \midrule \\[-1.3em]
    \begin{tikzpicture}[baseline={($(A.base)!.5!(B.base)$)}]
      \node[latent] (A) {$A$};
      \node[latent,below left =of A] (B) {$B$};
      \node[latent,below right=of A] (C) {$C$};
      \node[latent,right = 2 of A] (D) {$D$};
      \node[latent,below =of D] (E) {$E$};
      \edge {A} {B,C};
      \edge {D} {E};
    \end{tikzpicture} & $B +_{A+D} C +_{A+D} E$ \\
    \bottomrule
  \end{tabular}
\end{table}

To turn a Bayesian network into a Boolean algebra, run $f(\{ v \in V \mid
\mathrm{outdeg}(v) = 0 \})$, where $f$ works as follows on an arbitrary $A
\subseteq V$:
\begin{enumerate}
\item $S \gets \{ u \mid (u, v) \in E\text{, } v \in A \}$ (i.e., $S$ is the set
  of parents of the elements in $A$).
\item If $S = \emptyset$, return $\sum A$ (a coproduct).
\item Otherwise, return $\sum_{f(S)} A$ (a pushout).
\end{enumerate}
